Extreme climatic events with asymmetric tail-dependence affect population-extinction risk

### 2019-07-28 MainText.Rmd
discussion written for 1st draft

<!--Tuljapurkar approximation : linear version of age-structured model in stochastic environments-->
Though we found left tail dependence in vital rates (induced by ECEs) caused higher extinction risk than 
the right tail dependent ones, the difference was not remarkable (Fig. \ref{SM-fig_extrisk_agestr_C_SC_G_SG}) 
unless the tail dependence was at its extreme (Fig. \ref{fig_extrisk_agestr}). This finding could be explained 
by the approximation @tuljapurkar1990lecture made for the stochastic log(growth rate) in an iid environment.
The approximation considered that the growth rate depends on the most dominant eigen value of the mean projection matrix 
and the temporal variance of the log(population growth rate) caused by environmental stochasticity, which was based
on the linear covariance measure between any two matrix elements, not on their tail-dependence properties. Therefore,
according to Tuljapurkar's approximation, population growth rate in an iid environment 
remained same for our model (Eq. \ref{eq_age_str}) with left-tail or right-tail dependent noises.

<!--future study: worthy for PVA, IUCN red list
1. Identify which life stage or demographic process as management target?
2. Old study (Doak1994, Ferguson1995) showed including correlation btw vital rates for PVA makes a difference, 
   what if we include TD through copula?
3. multisite age str. model with dispersal, directional change in spatio-temporal variation for vital rates.
-->
Although we demonstrated that tail dependence between vital rates of an age structured model can
influence extinction risk, substantial work remains to determine the importance of this effect in the 
following scenarios. First, we demonstrated the effect of tail-dependence on extinction risk 
with a very simplistic age structured (two age stages) model where we considered the total semelperous 
population would go extinct if there would be only one female adult and one egg. But in reality population 
viability analysis (PVA) deals with a little bit higher threshold with several age, stage or size class and a variety 
of populations including semelperous and iteroperous [@morris2002quantitative]. Future studies should include 
tail-dependence effect on extinction risk 
for a multiple age-, size- and stage- structured model and estimate extinction risk 
for each age, size and stage class separately 
so that we can identify (via multisite-PVA) which life-stage or demographic process 
to target for population management. Earlier study by @doak1994modeling showed PVA models including correlation 
between different demographic rates indicated marked difference in predicting confidence interval for 
future population size than the model which did not include correlation. But when 
such correlation cannot be estimated directly, more generally where the
forms of statistical dependence among the model's parameters are unknown, making incorrect
assumptions about correlations and dependencies may lead to severe over-estimation 
or under-estimation of extinction risks [@ferson1995correlations]. We believe modeling 
dependencies with copula could help in this case. Second, we did not consider multiple sites of 
age-structured metapopulation which could be linked via dispersal. So, possible future investigation 
would be to explore the viability for the metapopulations if vital rates across sites 
showed specific tail dependence with spatially correlated extreme climatic events.

<!--What if TD with negative correlation?
e.g. some events that will kill adults will allow more successful recruitment of seeds or other propagules.
-->

<!--future study: more complex stage or age structured model, extinction risk how propagated through a food web-->

>>>>

<!--Tuljapurkar approximation : linear version of age-structured model in stochastic environments-->
We found left tail dependence in vital rates (induced by ECEs) caused higher extinction risk than 
the right tail dependent ones, but the difference was not remarkable (Fig. \ref{SM-fig_extrisk_agestr_C_SC_G_SG}) 
unless the tail dependence was at its extreme (Fig. \ref{fig_extrisk_agestr}). This finding could be explained 
by the approximation @tuljapurkar1990lecture made for the stochastic log(growth rate) in an iid environment.
The approximation considered that the growth rate depends on the most dominant eigen value of the mean projection matrix 
and the temporal variance of the log(population growth rate) caused by environmental stochasticity, which was based
on the linear covariance measure between any two matrix elements, not on their tail-dependence properties. Therefore,
according to Tuljapurkar's approximation, population growth rate in an iid environment 
remained same for our model (Eq. \ref{eq_age_str}) with left-tail or right-tail dependent noises.

<!--future study: worthy for PVA, IUCN red list
1. Identify which life stage or demographic process as management target?
2. Old study (Doak1994, Ferguson1995) showed including correlation btw vital rates for PVA makes a difference, 
   what if we include TD through copula? Multisite age str. model with dispersal, directional change
   in spatio-temporal variation for vital rates.
3. Tdep. with -ve correlation between vital rates?
4. extinction risk how propagated through a food web?
-->
Although we demonstrated that tail dependence between vital rates of an age structured model can
influence extinction risk, substantial work remains to determine the importance of this effect in the 
following scenarios. First, we demonstrated the effect of tail-dependence on extinction risk 
with a very simplistic age structured (two age stages) model where we considered the total semelperous 
population would go extinct if there would be only one female adult and one egg. But in reality population 
viability analysis (PVA) deals with a little bit higher threshold with several age, stage or size class and a variety 
of populations including semelperous and iteroperous [@morris2002quantitative]. Future studies should include 
tail-dependence effect on extinction risk 
for a multiple age-, size- and stage- structured model and estimate extinction risk 
for each age, size and stage class separately 
so that we can identify (via multisite-PVA) which life-stage or demographic process 
to target for population management. Earlier study by @doak1994modeling showed PVA models including correlation 
between different demographic rates indicated marked difference in predicting confidence interval for 
future population size than the model which did not include correlation. But when 
such correlation cannot be estimated directly, more generally where the
forms of statistical dependence among the model's parameters are unknown, making incorrect
assumptions about correlations and dependencies may lead to severe over-estimation 
or under-estimation of extinction risks [@ferson1995correlations]. We believe modeling 
dependencies with copula could help in this case. Second, we did not consider multiple sites of 
age-structured metapopulation which could be linked via dispersal. So, possible future investigation 
would be to explore the viability for the metapopulations if vital rates (e.g. fecundity or 
survival of different stage class) across sites showed specific tail dependence 
with spatially correlated extreme climatic events. Third, we considered tail dependence 
between vital rates when they were positively correlated, but what would be the extinction risk 
and viability of a population if demographic rates were negatively correlated with asymmetric tail dependence. 
This scenario could be considered for those extreme events that kill adults and allow more successful
recruitment of seeds or other propagules. Fourth, following the aforementioned idea, it would be worthy 
to model more complex prey-predator model where asymmetric tail dependent copula could be used to mimic
interspecific-dependence. An earlier study [@carnicer2011widespread] showed extreme events
(i.e., heat waves and droughts) increased mortality and defoliation in southern European forest and its 
dependent herbivorous insects and fungi through trophic linkage. ECEs can even truncate 
multiyear insect outbreak dynamics [@esper20061200]. We believe copula (may be a multidimensional one) has 
a lot more potential to estimate risk of cascading extinction in model food webs.
